---
title: "regplane3D: Plotting regression predictions in 3D"
author: "Denis Cohen & Nick Baumann"
date: "`r format(Sys.Date(), format='%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{regplane3D: Plotting regression predictions in 3D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(regplane3D)
```

# The regplane3D package

The `regplane3D` package is a convenience wrapper for Karline Soetaert's
[`plot3D`](https://cran.r-project.org/web/packages/plot3D/index.html) package.
`regplane3D` uses several `plot3D` functions to produce visually appealing
three-dimensional displays of regression estimates with confidence intervals.
For example, the package can be used to  plot conditional expected values 
of an outcome variable $Z$ over the joint distribution of two continuous
predictors, $X$ and $Y$, i.e., $\mathbb{E}(Z|X,Y)$.

`regplane3D` (development version 0.1.0) consists of the following functions:

1. `plane3D`: Plot a three-dimensional regression prediction with confidence intervals.
1. `twoplanes3D`: Plot a three-dimensional regression prediction with two planes, typically separated at a cut point in one of the two horizontal dimensions, with their respective confidence intervals.
1. `heatmap3D`: Auxiliary function for adding three-dimensional heatmaps to plots produced by either `plane3D` or `twoplanes3D`. These heatmaps show the joint frequency/density distribution of the model predictors represented on the horizontal axes of the plots.
1. `pretty_axis_inputs`: Auxiliary function for generating inputs for prediction and plotting to ensure that the grid lines of the perspective box and the lines of the grid lines of the regression planes match.

`regplane3D` is developed and maintained by
[Denis Cohen](https://twitter.com/denis_cohen) (author, creator) with the help
of [Nick Baumann](https://twitter.com/Nick_Baumann97) (contributor).

## Installation

To install the latest development version of `regplane3D` from GitHub, run:

```{r, eval = FALSE}
## devtools
if (!("devtools" %in% installed.packages()))
  install.packages("devtools")
library(devtools)

## regplane3D
if (!("regplane3D" %in% installed.packages()))
  devtools::install_github("denis-cohen/regplane3D")
library(regplane3D)
```

## Prerequisites

The use of `regplane3D` functions requires that users provide the following inputs:

1. A vector containing a sequence of values for the first predictor, $X$.
1. A vector containing a sequence of values for the second predictor, $Y$.
1. A matrix containing the expected values of $Z$, or an array containing the expected values as well as their lower and upper confidence interval bounds, for all combinations of the specified values for $X$ and $Y$.
1. *Optional*: A matrix containing the discretized joint density or joint frequency of $X$ and $Y$.

We illustrate how these inputs can be generated in some applied examples below.

# Using regplane3D functions

We will illustrate the use of `regplane3D`'s functions using the package's
internal data set `us`, a small data set containing information on incumbent
vote shares, approval ratings, and economic growth rates in
US Presidential Elections 1948-2004.
For a documentation of the data, see `?regplane3D::us`.

## plane3D()

A staple of introductory statistics classes is the notion that "we no longer
deal with a line in two dimensions, but with a plane in three dimensions" once
we extend the simple bivariate regression model to include an additional
predictor. Our first example uses the `plane3D()` function to illustrate
this in the context of an OLS regression of incumbent vote shares in US
Presidential Elections on incumbent approval ratings and economic growth.

### Definition of axis inputs

When using `regplane3D` plotting functions, it is recommended that users use
`regplane3D::pretty_axis_inputs()` for defining axis inputs that should be used
for *both* prediction *and* plotting.
The reason for this requires some explanation. `regplane3D` plotting functions
use `plot3D::perspbox()` to generate the perspective box inside which the
plots are drawn. `plot3D::perspbox()` in turn depends on `graphics::persp()`,
which uses the base R function `pretty()` to determine the number of ticks
and reference lines of the perspective box. 

As a result of these dependencies, users have limited control over the exact
number and placement of ticks and reference lines. For instance, if users were to
provide a variable ranging from 1.89 to 24.31 and request four ticks,
this suggestion will be overridden with a rounded value range from 0 to 25 and
a total of six ticks in integer steps of 5:

```{r pretty-example}
pretty(c(1.89, 24.31), n = 4)
```

Therefore, we recommend that users anticipate this particularity early on and
define their axis inputs such that predictions and their visualization
eventually conform to the grid lines of the perspective box. The function 
`pretty_axis_inputs()` performs these tasks. It rounds value ranges to a
custom base and provides the number and positions of the grid lines in the
perspective box. These can then be used *before* the plot is generated to
predict the regression plane at the corresponding values. To provide for
smoother curves in plots involving curvilinear planes, the option `multiply`
ensure that values of the plane are not only calculated at the intersections of
the grid but at finer gradations.

The example below illustrates the functionality. The function extends the
range of the variable `us$growth` (`r min(us$growth)`, `r max(us$growth)`) to a
`base` of 2, such that the coarsened range is (-4, 6). We suggest that this 
range be split into 7 equally spaced intervals. This is rejected by the function,
as such a division would not yield pretty values.
Instead, the function returns
`nlines = 6`, which means that the lines should be drawn at the reported
`linevals` of -4, -2, 0, 2, 4, and 6.
When plotting non-linear relationships, grid lines evaluated at such few values
may look a little jerky. To obtain smoother predictions, we can compute our
expected values not just at these coarse values but also at finer gradations
in between. To accomplish this, we can for instance specify `multiply = 4`,
which means that the `linevals` sequence in steps of two will be divided into
a finer sequence in steps of 0.5, which is returned as `seq`.

```{r growth-pretty}
## Find range of variable
growth_range <- range(us$growth)
growth_range

## Determine axis inputs
growth_axis <- pretty_axis_inputs(
  axis_range = growth_range,
  base = 2,
  nlines_suggest = 7L,
  multiply = 4
)
growth_axis
```

### Estimation and prediction

To obtain the required inputs, we first run a linear regression model of the form
$\texttt{vote} = \beta_1 + \beta_2 \texttt{growth} + \beta_3 \texttt{approval} +\epsilon$
and save the estimation results to an object named `mod`.

```{r use-plane3D-step1}
## ---- Estimation ----
mod <- lm(vote ~ growth + approval, dat = us)
```

We then use `pretty_axis_inputs()` to define the inputs for both axes. This
gives us the sequence of values for each `growth` and `approval`. 

```{r apply-pretty}
## ---- Axis inputs ----
## Growth
growth_axis <- pretty_axis_inputs(
  axis_range = range(us$growth),
  base = 2,
  nlines_suggest = 6L,
  multiply = 1
)

## Approval
approval_axis <- pretty_axis_inputs(
  axis_range = range(us$approval),
  base = 10,
  nlines_suggest = 6L,
  multiply = 1
)
```

For every combination of the values of these two sequences, we subsequently
calculate the expected value and the lower and upper bounds of 
its 95% confidence interval using the `predict.lm()` function with option
`se.fit = TRUE`.
At each iteration of the nested loop, expected values are temporarily stored in
`pred_tmp$fit` and standard errors are temporarily stored in `pred_tmp$se.fit`.
We can extract the expected value and calculate the lower and upper bounds of
the 95% confidence interval at each iteration using
`pred_tmp$fit + qnorm(.025) * pred_tmp$se.fit` and
`pred_tmp$fit + qnorm(.975) * pred_tmp$se.fit`, respectively.
We subsequently store the estimate of a given iteration in the appropriate cell
of the array `pred`.
The array is of dimensions `dim = c(length(growth_axis$seq), length(approval_axis$seq), 3L)`.
The first dimension represents the values of `growth_axis$seq`, the second dimension
represents the values of `approval_axis$seq`, and the third dimension represents the
point estimates, lower confidence bounds, and upper confidence bounds.

```{r use-plane3D-step2}
## ---- Prediction ----
pred <-
  array(NA, dim = c(length(growth_axis$seq), length(approval_axis$seq), 3L))
for (growth in seq_along(growth_axis$seq)) {
  for (approval in seq_along(approval_axis$seq)) {
    pred_tmp <- predict.lm(
      mod,
      newdata = data.frame(growth = growth_axis$seq[growth],
                           approval = approval_axis$seq[approval]),
      se.fit = TRUE
    )
    pred[growth, approval, ] <- c(
      pred_tmp$fit,
      pred_tmp$fit + qnorm(.025) * pred_tmp$se.fit,
      pred_tmp$fit + qnorm(.975) * pred_tmp$se.fit
    )
  }
}
```

*Note:* The prediction step can (and should) be adopted to fit the requirements
of a given
empirical application. For instance, the calculation of expected values from
generalized linear models requires the specification of scenarios for the
covariate values, the application of an inverse link function, and the use of
bootstrapping or parameter simulation for the construction of confidence
intervals (though the last step may be skipped in favor of analytical
confidence intervals based on normal approximation if the sampling distribution
of the quantity of interest is approximately normal). For more information on
the calculation of quantities of interest in generalized linear models, see
the [Further reading](#further-reading) section.

### Plotting

Using these estimates, we can then plot our regression plane using
`plane3D()`.
We pass the inputs `z = pred`, `x = growth_axis$seq`, and
`y = approval_axis$seq` to the
function, which contain all required information to plot the regression plane.
The point estimate of the regression line is plotted by default.
Confidence intervals are added per the option `cis = TRUE`.
For additional options, see `?plane3D`.

```{r use-plane3D-step3, fig.width=6.5, fig.height=5, out.width='80%', fig.align='center', dpi=300}
## ---- Plot ----
par(mar = c(2.1, 2.1, 4.1, 0.1))
plane3D(
  z = pred,
  x = growth_axis$seq,
  y = approval_axis$seq,
  zlab = "Predicted Vote Share",
  xlab = "Economic Growth",
  ylab = "Approval Rating",
  zlim = c(35, 70),
  xlim = growth_axis$range,
  ylim = approval_axis$range,
  cis = TRUE,
  xnlines = growth_axis$nlines,
  ynlines = approval_axis$nlines,
  main = "Incumbent Vote Shares, Economic \n Growth, and Approval Ratings",
  theta = -45,
  phi = 9
)
```

### Extensions

The plot can be extended using `regplane3D::heatmap3D()`, which adds a
three-dimensional histogram to show the joint frequency or density distribution
of `growth` and `approval`.
Toward this end, we must first compute a matrix of the joint frequency along
discrete intervals of the two continuous predictors. For appealing visuals,
it is recommended that the partition of the discrete intervals corresponds
to the grid lines of the main plot; i.e., approval rating should be discretized
from 30 to 80 in intervals of 10 and economic growth should be discretized 
from -4 to 6 in steps of 2:

```{r use-plane3D-step4}
## Heatmap values
growth_cat <- cut(us$growth, breaks = growth_axis$linevals)
approval_cat <- cut(us$approval, breaks = approval_axis$linevals)
joint_frequency <- table(growth_cat, approval_cat)
```

We can then add the three-dimensional heatmap by adding the option
`heatmap = joint_frequency` to our `plane3D()` command:

```{r use-plane3D-step5, fig.width=6.5, fig.height=5, out.width='80%', fig.align='center', dpi=300}
## ---- Plot ----
par(mar = c(2.1, 2.1, 4.1, 0.1))
par(mar = c(2.1, 2.1, 4.1, 0.1))
plane3D(
  z = pred,
  x = growth_axis$seq,
  y = approval_axis$seq,
  zlab = "Predicted Vote Share",
  xlab = "Economic Growth",
  ylab = "Approval Rating",
  zlim = c(35, 70),
  xlim = growth_axis$range,
  ylim = approval_axis$range,
  cis = TRUE,
  xnlines = growth_axis$nlines,
  ynlines = approval_axis$nlines,
  main = "Incumbent Vote Shares, Economic \n Growth, and Approval Ratings",
  theta = -45,
  phi = 9,
  heatmap = joint_frequency
)
```

As the `regplane3D` package is a convenience wrapper for the `plot3D` package,
plots produced by `regplane3D` plotting functions can be supplemented with
output from `plot3D` functions (using the option `add = TRUE`).
For instance, we can add the observed values of the outcome variable 
using `plot3D::points3D()` and add text labels using `plot3D::text3D()`.

```{r use-regplane3D-step6, fig.width=6.5, fig.height=5, out.width='80%', fig.align='center', dpi=300}
## ---- Plot ----
par(mar = c(2.1, 2.1, 4.1, 0.1))
plane3D(
  z = pred,
  x = growth_axis$seq,
  y = approval_axis$seq,
  zlab = "Predicted Vote Share",
  xlab = "Economic Growth",
  ylab = "Approval Rating",
  zlim = c(35, 70),
  xlim = growth_axis$range,
  ylim = approval_axis$range,
  cis = TRUE,
  xnlines = growth_axis$nlines,
  ynlines = approval_axis$nlines,
  main = "Incumbent Vote Shares, Economic \n Growth, and Approval Ratings",
  theta = -45,
  phi = 9,
  heatmap = joint_frequency
)
plot3D::points3D(
  z = us$vote,
  x = us$growth,
  y = us$approval,
  add = TRUE,
  col = adjustcolor("black", alpha.f = .3),
  pch = 19
)
plot3D::text3D(
  z = us$vote + 2.5,
  x = us$growth,
  y = us$approval,
  labels = us$incumbent,
  add = TRUE,
  cex = 0.6
)
```

## twoplanes3D()

The `twoplanes3D()` function extends the functionality of `plane3D()` to
accommodate two separate planes. These are typically required when the
model prediction is distinct to specific value ranges separated by a cut point
in one of the two horizontal dimensions, akin to a discontinuity or binary
spline.

We showcase the function by replicating the empirical example introduced
above, now with distinct predictions for incumbent presidents
with above-average and below-average approval ratings, respectively.

### Axis inputs, estimation and prediction

Axis inputs, estimation and prediction are now slightly more intricate than in 
the previous example. First, we interact both model
predictors with the binary indicator `approval_above_mean`.
We then define the axis input for the cut axis (i.e., `centered_approval`) from
its minimum value  up to the `cut_point`of 0. 
We  store the expected values and confidence bounds across the values of
`growth_axis$range`
and approval values ranging from `min(centered_approval)` up to the cut point of
0 in `pred[, , 1, ]` for the prediction with below-average approval ratings. 
Analogously, we store this information for the prediction with above-average
approval ratings in `pred[, , 2, ]`, where the values of `centered_approval`
now range from the cut point of up to `abs(min(centered_approval))` to provide
for a symmetrical value range and display.

<details>
<summary> R code: Estimation and prediction</summary>
```{r use-twoplanes3D-step1}
## ---- Estimation ----
mod <-
  lm(vote ~
       growth +
       centered_approval +
       approval_above_mean +
       growth:approval_above_mean +
       centered_approval:approval_above_mean,
     dat = us)

## ---- Axis inputs ----
## Cut point
cut_point <- 0
approval_above_mean_vals <- c(0, 1)

## Growth
growth_axis <- pretty_axis_inputs(
  axis_range = range(us$growth),
  base = 2,
  nlines_suggest = 6L,
  multiply = 1
)

## Approval
approval_axis <- pretty_axis_inputs(
  axis_range = c(min(us$centered_approval), cut_point),
  base = 10,
  nlines_suggest = 3L,
  multiply = 1
)

## ---- Prediction ----
pred <-
  array(NA, dim = c(length(growth_axis$seq), length(approval_axis$seq), 2L, 3L))
for (growth in seq_along(growth_axis$seq)) {
  for (centered_approval in seq_along(approval_axis$seq)) {
    for (approval_above_mean in seq_along(approval_above_mean_vals)) {
      pred_tmp <- predict.lm(
        mod,
        newdata = data.frame(
          growth = growth_axis$seq[growth],
          centered_approval = approval_axis$seq[centered_approval] -
            min(approval_axis$seq) *
            approval_above_mean_vals[approval_above_mean],
          approval_above_mean = approval_above_mean_vals[approval_above_mean]
        ),
        se.fit = TRUE
      )

      pred[growth, centered_approval, approval_above_mean,] <-
        c(
          pred_tmp$fit,
          pred_tmp$fit + qnorm(.025) * pred_tmp$se.fit,
          pred_tmp$fit + qnorm(.975) * pred_tmp$se.fit
        )
    }
  }
}
```
</details>
</br>

### Plotting

Plotting with `twoplanes3D()` works the same way as with `plane3D()`, except
we now must provide the $x$, $y$ and $z$ values separately for both planes.
For this, we use the inputs `x` and `x2`, `y` and `y2`, as well as `z` and `z2`.
Extensions can be added in the same way as with `plane3D()`.

```{r use-twoplanes3D-step2, fig.width=6.5, fig.height=5, out.width='80%', fig.align='center', dpi=300}
## ---- Plot ----
par(mar = c(2.1, 2.1, 4.1, 0.1))
twoplanes3D(
  z = pred[, , 1,],
  x = growth_axis$seq,
  y = approval_axis$seq,
  z2 = pred[, , 2,],
  x2 = growth_axis$seq,
  y2 = approval_axis$seq - min(approval_axis$seq),
  zlim = c(35, 70),
  xlim = growth_axis$range,
  ylim = c(min(approval_axis$seq),-min(approval_axis$seq)),
  zlab = "Predicted Vote Share",
  xlab = "Economic Growth",
  ylab = "Approval Rating \n Above & Below Average",
  cis = TRUE,
  xnlines = growth_axis$nlines,
  ynlines = approval_axis$nlines,
  main = "Incumbent Vote Shares, Economic \n Growth, and Approval Ratings",
  theta = -55,
  phi = 9
)
```
